vegan_analysis

### 2024-01-10 scripts/4_vegan_clean.qmd
began revisions round 2, cleaned 4_vegan to be easier to work with (less defunct extra code)

```{r packages}
library(vegan)
library(tidyverse)
library(cols4all)
library(terra)
library(sf)
library(patchwork)
# library(ggradar)
source(here::here("scripts", "get_keys.R"))
```

@@

```{r data-prep}
bec_sample_loc <- here::here("data", "samples", "vlce_extracted.shp")

@@

bec_normal <- bec_df %>%
  mutate(across(basal_area:VarDHI,  .fns = ~ normal(.x)))

```{r global-model}
to_rda <- bec_normal %>%
  select(-elev_mean,
         -percentage_first_returns_above_mean,
         -elev_stddev,
         -loreys_height)

forest_types <- bec_normal %>%
  pull(category) %>%
  unique()

# to_rda <- bec_normal %>% 
#   group_by(category) %>% 
#   slice_sample(n = 3000)  %>% 
#   filter(category == "Mixed Wood") %>% 
#   ungroup()

dhi_tab <- to_rda %>%
  select(ends_with("DHI"))

primary <- to_rda %>%
  select(elev_p95,
         elev_cv,
         # elev_stddev,
         percentage_first_returns_above_2m)

modelled <- to_rda %>%
  select(basal_area,
         total_biomass,
         gross_stem_volume)

rda_out <- 
  rda(dhi_tab ~ elev_p95 + percentage_first_returns_above_2m + elev_cv + # simple/raw/primary
      basal_area + gross_stem_volume + total_biomass, # modelled
      data = to_rda)

@@

# scores(rda_out, choices = c(1:3))$species
# scores(rda_out, choices = c(1:3))$biplot
# not actually the loadings, need to get those from correlations

rcorr <- summary(rda_out)$sites %>% 
  as.data.frame() %>% 
  select(starts_with("RDA")) %>% 
  bind_cols(to_rda) %>%
  select(-bec, -category) %>% 
  as.matrix() %>%
  Hmisc::rcorr()

@@

all_data_loadings <- left_join(rvals, pvals) %>%
  filter(axis != "RDA3") %>%
  mutate(sig = cut(axis_p, 
                   breaks = c(0, 0.001, 0.01, 0.05, .10, 1),
                   labels = c("***", "**", "*", ".", "NS"),
                   right = F))

for_radar_loads <- all_data_loadings %>%
  filter(!endsWith(var, "DHI")) %>%
  mutate(axis_loading = ifelse(sig == "NS", 0, abs(axis_loading))) %>%
  select(var, axis, axis_loading) %>%
  pivot_wider(names_from = var, values_from = axis_loading) %>%
  mutate(cluster = paste(axis, "- British Columbia")) %>%
  select(-axis) %>%
  relocate(cluster)

@@

Metrics derived solely from the point cloud explain 10.6% of the variation in the DHIs, while metrics modelled based on the point cloud explain 3.4% of the variation in the DHIs. 85.9% of the variation is unexplained by structural attributes.

@@

This model shows very little explanatory power. (13.9%)

@@

The model is statistically significant (P \< 0.001). All of the axes are significant, although this may be related to the large sample size.

# # vlce sampling regime to get the samples up
# switched away from this to a single double stratified sample
# vlce_samples <- read_sf(here::here("data", "samples", "vlce_extracted.shp"))
# 
# vlce_df <- vlce_samples %>%
#   st_drop_geometry() %>%
#   select(-category_1, -category_2) %>%
#   rename(category = strata) %>%
#   relocate(bec) %>%
#   drop_na() %>%
#   ungroup()
# 
# names(vlce_df) <- names_manual
# 
# vlce_normal <- vlce_df %>%
#   mutate(across(basal_area:VarDHI,  .fns = ~ normal(.x))) 
# 
# to_nest <- vlce_normal %>%
#   select(-percentage_first_returns_above_mean,
#          -loreys_height,
#          -elev_stddev,
#          -elev_mean)
# 
# 
# 
# 
# class_nested <- to_nest %>%
#   group_by(category) %>%
#   slice_sample(n = 3000) %>%
#   nest() %>% 
#   ungroup() %>%
#   rename(strata = category)

strata_nested <- bind_rows(bec_nested, class_nested)
```

@@

bec_normal %>%
  pivot_longer(cols = ends_with("DHI"),
               names_to = "dhi",
               values_to = "dhi_val") %>%
  pivot_longer(
    cols = basal_area:total_biomass,
    names_to = "struct",
    values_to = "struct_val"
  ) %>%
  ggplot(aes(x = struct_val, y = dhi_val)) +
  geom_hex() +
  facet_grid(rows = vars(dhi),
             cols = vars(struct),
             scales = "free") +
  labs(x = "Structure Value (Normalized)",
       y = "DHI Value (Normalized)")

strata_nested_long <- strata_nested %>%
  unnest(cols = data) %>%
  pivot_longer(cols = ends_with("DHI"),
               names_to = "dhi",
               values_to = "dhi_val") %>%
  pivot_longer(
    cols = basal_area:total_biomass,
    names_to = "struct",
    values_to = "struct_val"
  ) %>%
  group_by(strata) %>%
  group_split()

test <- function(df) {
  strata <- df %>%
    pull(strata) %>%
    unique()

  plot <- df %>% ggplot(aes(x = struct_val, y = dhi_val)) +
    geom_hex() +
    facet_grid(rows = vars(dhi),
               cols = vars(struct),
               scales = "free") +
    labs(x = "Structure Value (Normalized)",
         y = "DHI Value (Normalized)")

  ggsave(here::here("outputs", "linear", glue::glue("{strata}.png")))
}

map(strata_nested_long, test)
```

@@

```{r fracts-out}

fracts_out <- outs %>%
  mutate(indfracts = map(var, .f = function(x) {
    x$part$indfract %>%
      mutate(type = c("primary", "modelled", "overlap", "residual")) %>%
      select(type, Adj.R.squared) %>%
      pivot_wider(names_from = type, values_from = Adj.R.squared)
    }
    )) %>%
  unnest(indfracts) %>%
  arrange(desc(residual)) %>%
  select(strata, primary, modelled, overlap) %>%
  pivot_longer(c(primary, modelled, overlap)) %>%
  mutate(value = ifelse(value > 0, value, 0)) %>%
  pivot_wider()

@@

@@

```{r raincloud-plot}
library(ggdist)
fracts_out %>%
  mutate(total = primary + modelled + overlap) %>%
  #left_join(keys$bec, by = c("bec" = "zone")) %>%
  pivot_longer(-strata) %>%
  mutate(name = str_to_title(name)) %>%
  mutate(value = ifelse(value < 0, 0, value),
         class = ifelse(strata %in% forest_types, 
                        "Forest Types", 
                        "BEC Zones")) %>%
  ggplot(aes(x = value, y = name)) +
  stat_halfeye(
    # adjust bandwidth
    adjust = .5,
    # move to the right
    justification = -0.2,
    # remove the slub interval
    .width = 0,
    point_colour = NA
    ) +
  geom_boxplot(
    width = 0.12,
    # removing outliers
    outlier.color = NA,
    alpha = 0.5
  ) +
  stat_dots(
    # ploting on left side
    side = "left",
    # adjusting position
    justification = 1.1,
    # adjust grouping (binning) of observations
    binwidth = 0.0025,
    #position = "stack",
    # aes(fill = group, col = group),
    alpha = 0.5,
    aes(shape = class)
  ) +
  labs(y = NULL,
       x = "% Variance Explained",
       fill = "BEC Group",
       col = "BEC Group") +
  scale_x_continuous(labels = scales::label_percent()) +
  theme_bw() +
  theme(panel.grid = element_blank(),
        legend.position = c(0.8, .2))

```{r path-setup}
path_base <- tibble(x1 = c(0,
             0,
             0,
             0,
             1),
       y1 = c(1,
              -1,
              1,
              -1,
              0),
       x2 = c(1.85,
              1.85,
              .85,
              .85,
              1.85),
       y2 = c(.15,
              -.15,
              .15,
              -.15,
              0)) %>%
  ggplot(aes(x = x1, y = y1)) +
  geom_segment(aes(xend = x2, yend = y2), arrow = arrow()) +
  geom_point(data = tibble(x1 = c(.85, .85), 
                           y1 = c(-.5, .5)),
             size = 60,
             shape = 21,
             col = "grey") +
  # geom_text(aes(x = 0, y = -1.15), label = "Modelled\nAttributes") +
  # geom_text(aes(x = 0, y = 1.15), label = "primary\nAttributes") +
  # geom_text(aes(x = .65, y = 0), label = "Overlap") +
  # geom_text(aes(x = 2, y = 0), label = "Dynamic\nHabitat\nIndices") +
  theme_void() +
  scale_x_continuous(limits = c(-0.1, 2.1))

ggsave(here::here("outputs", "path_base.png"), path_base)

path_r2_plot <- tibble(type = c("primary", "modelled", "overlap"),
       x1 = c(0.85, 0.85, 0.85),
       y1 = c(0.75, -0.75, 0))
```

```{r path-global}
var_oi <- c("VarDHI", "MinDHI", "CumDHI")
#var_oi <- c("VarDHI", "MinDHI")

var_out <- varpart(dhi_tab %>%
                     select(any_of(var_oi)),
                   primary, modelled)

# plot(var_out, Xnames = c("Primary", "Modelled"))

path_r2 <- var_out$part$indfract %>%
      mutate(type = c("primary", "modelled", "overlap", "residual")) %>%
      select(type, Adj.R.squared) %>%
  tibble()

path_r2_clean <- path_r2_plot %>%
  left_join(path_r2) %>%
  mutate(r2_label = paste0(round(Adj.R.squared * 100, 2), "%"))

path_base +
  geom_text(data = path_r2_clean, aes(label = r2_label))

ggsave(here::here("outputs", "path_venn.png"),
        height = 3.5, width = 10)
```

@@

```{r cum-min-hexplot}
bec_df %>% 
  select(bec, CumDHI, MinDHI) %>%
  left_join(keys$bec, by = c("bec" = "zone")) %>%
  ggplot(aes(x = CumDHI, y = MinDHI)) +
  geom_hex() +
  facet_wrap(~bec)
```

@@

axis_varexp %>%
  filter(anova_sig != "NS") %>%
  mutate(category = ifelse(strata %in% forest_types, "Forest Types", "BEC Zones")) %>%
  # left_join(keys$bec, by = c("bec" = "zone")) %>%
  ggplot(aes(x = axis, y = var_exp, col = category, shape = category)) +
  geom_jitter(height = 0, width = 0.05) +
  theme(legend.position = "bottom") +
  labs(x = NULL,
       y = "% of Variance Explained",
       col = NULL,
       shape = NULL) +
  scale_y_continuous(labels = scales::label_percent())
# no secondary axis in alpine & coast.

@@

```{r varexp-coniferous}
rda_varexp <- outs %>%
  mutate(varexp = map(rda, .f = \(rda) {
    rda_sum <- summary(rda)
    var_exp <- rda_sum$constr.chi / rda_sum$tot.chi
  })) %>%
  unnest(varexp) %>%
  arrange(desc(varexp)) %>%
  select(strata, varexp)

rda_varexp %>%
  arrange(strata %in% forest_types,
          desc(varexp))

source(here::here("scripts", "get_bec_proportion_vlce.R"))

prop_con <- bec_normal %>% 
  select(bec, category) %>% 
  group_by(bec) %>% 
  summarize(per_sample_for = mean(category == "Coniferous")) %>% 
  left_join(per_cons %>% 
              select(-value)) %>% 
  mutate(for_dif = per_sample_for - per_con) %>%
  arrange(desc(abs(for_dif))) %>%
  select(-for_dif) %>%
  pivot_longer(starts_with("per"))

con_var_plot <- prop_con %>%
  filter(name == "per_con") %>%
  left_join(rda_varexp, by = c("bec" = "strata")) %>% 
  left_join(keys$bec, by = c("bec" = "zone")) %>%
  mutate(name = ifelse(name == "per_con", 
                       "Global Percent Coniferous", 
                       "Sample Percent Coniferous")) %>%
  ggplot(aes(x = value, y = varexp)) +
  geom_point(aes(col = group)) +
  geom_smooth(method = "lm") +
  scale_x_continuous(labels = scales::label_percent()) +
  scale_y_continuous(labels = scales::label_percent()) +
  labs(x = "Percent Coniferous",
       y = "Proportion of variance explained",
       col = NULL) +
  theme(legend.position = c(0.1, 0.85))

ggsave(here::here("outputs", "con_var.png"), plot = con_var_plot)
```

@@

sig_loads %>%
  filter(anova_sig != "NS") %>%
  filter(endsWith(var, "DHI"),
         axis == "RDA2") %>%
  arrange(strata) %>%
  select(strata, var, axis_loading) %>%
  pivot_wider(names_from = var,
              values_from = axis_loading)
# when a second axis is present, it has a weak loading on the cumulative DHI, as compared to the other DHIs
# while the majoirty of the variability is contained within all three DHIs together, the second axis, when present, is more linked to the minmimum and variation DHI.

sig_loads %>%
  select(strata, var, axis, axis_loading, loading_sig) %>%
  filter(endsWith(var, "DHI"),
         loading_sig != "NS") %>%
  ggplot(aes(x = var, y = abs(axis_loading), group = strata)) +
  geom_point(aes(col = strata)) +
  geom_line(aes(col = strata), alpha = 0.5, lty = "dashed") +
  geom_hline(yintercept = 0, col = "red", alpha = 0.2) +
  facet_wrap(~axis) +
  theme_bw() +
  theme(panel.grid = element_blank()) +
  labs(x = NULL,
       y = "Axis Loading")

sig_loads %>%
  select(strata, var, axis, axis_loading) %>%
  filter(!endsWith(var, "DHI")) %>%
  left_join(keys$continuous, by = c("var" = "variable")) %>%
  ggplot(aes(x = var_long, y = abs(axis_loading), group = strata)) +
  geom_point(aes(col = strata)) +
  geom_line(aes(col = strata), alpha = 0.5, lty = "dashed") +
  geom_hline(yintercept = 0, col = "red", alpha = 0.2) +
  facet_wrap(~axis) +
  theme(axis.text.x = element_text(angle = 45,
                                   hjust = 1)) +
  labs(x = NULL,
       y = "Axis Loading (Absolute Value)")

sig_loads %>%
  filter(!endsWith(var, "DHI"),
         loading_sig != "NS") %>%
  group_by(strata, axis) %>%
  mutate(loading_rank = data.table::frankv(abs(axis_loading), 
                                           order = -1)) %>%
  ggplot(aes(x = fct_reorder(var, loading_rank),
             y = loading_rank, col = strata)) +
  geom_jitter(height = 0) +
  facet_wrap(~axis) +
  theme(axis.text.x = element_text(angle = 45, 
                                   hjust = 1)) +
  labs(x = NULL,
       y = "Axis Loading Rank")

# need to remove the correlations for variables that arent actually involved
# percentage_first_returns_above_mean, elev_mean, elev_stddev
out_loadings %>% count(var)

@@

radar chart based on groups, not hierarchical clustering

```{r radar}
gg_color_hue <- function(n) {
  hues = seq(15, 375, length = n + 1)
  hcl(h = hues, l = 65, c = 100)[1:n]
}

n = 8
cols = c4a(palette = "rainbow", n = 8)

radar_loads <- clean_loads %>%
  pivot_wider(names_from = var, 
              values_from = axis_loading) %>%
  left_join(keys$bec, by = c("strata" = "zone")) %>%
  mutate(group = ifelse(is.na(group), "Forest Types", group),
         group = ifelse(strata == "British Columbia", "British Columbia", group),
         group = ifelse(axis == "RDA2", "RDA2", group)) %>%
  group_by(group) %>%
  summarize(across(CumDHI:elev_p95, \(x) mean(x, na.rm = T))) %>%
  mutate(group = fct_relevel(group, "RDA2", after = Inf),
         group = fct_relevel(group, "Forest Types", after = 1))

radar_labels <- radar_loads %>%
  select(-group, -ends_with("DHI")) %>%
  colnames() %>%
  tibble(variable = .) %>%
  left_join(keys$continuous) %>%
  pull(var_long)

in_radar <- radar_loads %>%
  select(!ends_with("DHI")) %>%
  replace(is.na(.), 0) %>%
  ggradar(axis.labels = radar_labels,
          legend.title = "Cluster",
          values.radar = c("", "", 1),
          base.size = 1,
          background.circle.transparency = 0,
          group.point.size = 2,
          group.colours = cols,
          axis.label.size = 3,
          plot.extent.x.sf = 1.5,
          group.line.width = 1) +
  theme(legend.position = "none")

out_radar <- radar_loads %>%
  select(group, ends_with("DHI")) %>%
  replace(is.na(.), 0) %>%

  ggradar(legend.title = "Cluster",
          values.radar = c("", "", 1),
          base.size = 1,
          background.circle.transparency = 0,
          group.point.size = 2,
          group.colours = cols,
          axis.label.size = 3,
          plot.extent.x.sf = 1.5,
          group.line.width = 1) +
  theme(legend.position = "none")

boxplots <- clean_loads %>% 
  left_join(keys$continuous, by = c("var" = "variable")) %>%
  mutate(var_long = ifelse(is.na(var_long), var, var_long)) %>%
  select(!(divide_by:unit)) %>%
  left_join(keys$bec, by = c("strata" = "zone")) %>%
  mutate(group = ifelse(is.na(group), "Forest Types", group),
         group = ifelse(axis == "RDA2", "RDA2", group),
         group = ifelse(strata == "British Columbia", "British Columbia", group),
         inout = ifelse(endsWith(var, "DHI"), "Response", "Predictor")) %>%
  mutate(group = fct_relevel(group, "RDA2", after = Inf),
         group = fct_relevel(group, "Forest Types", after = 1)) %>%
  ggplot(aes(x = abs(axis_loading), y = fct_rev(fct_reorder(var_long, abs(axis_loading))))) +
  geom_boxplot(outlier.colour = "#00000000", width = .5) +
  geom_jitter(height = 0.1, aes(col = group, shape = group)) +
  facet_grid(cols = vars(axis),
             rows = vars(inout),
             scales = "free_y",
             space = "free_y") +
  labs(x = "Axis Loading (absolute)",
       y = NULL,
       col = NULL,
       shape = NULL) +
  theme(panel.spacing.y = unit(0, "lines"),
        legend.position = "right") +
  scale_shape_manual(values = c(17, 15, 16, 16, 16, 16, 16, 16)) +
  scale_colour_manual(values = cols)

radar_boxplot <- ((in_radar / out_radar) | boxplots)  +
  plot_annotation(tag_levels = "A") +
  plot_layout(guides = "collect") &
  theme(plot.tag = element_text(size = 12))

ggsave(here::here("outputs", "radar_boxplot.png"),
       plot = radar_boxplot,
       device = "png",
       height = 6,
       width = 8)
```

@@

  write_lines(strata_drawio, here::here(drawio_dir, paste0(s, ".drawio")))
}
```

```{r complexity-iqr}
strata_nested %>% 
  unnest(data) %>% 
  ggplot(aes(x = fct_reorder(strata, elev_cv, IQR), y = elev_cv)) + 
  geom_boxplot() + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

@@

bec_tab <- keys$bec %>%
  left_join(bec_clim, by = c("zone" = "bec")) %>%
  select(-order) %>%
  arrange(group, zone) %>%
  relocate(group) %>%
  mutate(across(ppt:tmin, \(x) round(x, digits = 1)))

```{r}
bec_normal %>%
  select(-elev_stddev,-loreys_height,-percentage_first_returns_above_mean,-elev_mean) %>%
  left_join(keys$bec, by = c("bec" = "zone")) %>%
  pivot_longer(basal_area:VarDHI) %>%
  ggplot(aes(x = value, y = bec, fill = group)) +
  geom_boxplot() +
  facet_wrap( ~ name)
```

# new radar loads

```{r}
clean_loads %>%
  pivot_wider(names_from = var, 
              values_from = axis_loading) %>%
  left_join(keys$bec, by = c("strata" = "zone")) %>%
  mutate(group = ifelse(is.na(group), strata, group),
         group = ifelse(strata == "British Columbia", "British Columbia", group)) %>%
  group_by(group, axis) %>%
  summarize(across(CumDHI:elev_p95, \(x) mean(x, na.rm = T))) %>%
  pivot_longer(CumDHI:elev_p95) %>%
  mutate(value = ifelse(is.na(value), 0, value)) %>%
  left_join(keys$continuous, by = c("name" = "variable")) %>%
  mutate(var_long = ifelse(is.na(var_long), name, var_long),
         order = endsWith(var_long, "DHI")) %>%
  ggplot(aes(x = fct_reorder(var_long, value, .desc = T), y = value)) +
  geom_col() +
  #coord_polar() +
  coord_flip() +
  facet_grid(cols = vars(group), rows = vars(axis)) +
  labs(x = "Variable",
       y = "Axis Loading (absolute)")

clean_loads %>%
  left_join(keys$bec, by = c("strata" = "zone")) %>%
  mutate(group = ifelse(is.na(group), strata, group),
         group = ifelse(strata == "British Columbia", "British Columbia", group)) %>%
  ggplot(aes(x = var, y = axis_loading, col = group)) +
  geom_line()
```

```{r}
clean_loads %>%
  left_join(keys$bec, by = c("strata" = "zone")) %>%
  mutate(group = ifelse(is.na(group), strata, group),
         group = ifelse(strata == "British Columbia", "British Columbia", group)) %>%
  pivot_wider(names_from = var, values_from = axis_loading) %>%
  select(-zone_nm, -order) %>%
  replace(is.na(.), 0) %>%
  GGally::ggparcoord(columns = 4:12,
                     groupColumn = "group",
                     scale = "globalminmax") +
  facet_wrap(~axis) +
  coord_flip()
```

# parallel coordinate plots

```{r}
data <- clean_loads_abs %>%
  left_join(keys$bec %>%
              select(strata = zone, group)) %>%
  left_join(keys$continuous %>%
              select(var = variable, var_long)) %>%
  mutate(group = ifelse(is.na(group), strata, group),
         var_long = ifelse(is.na(var_long), var, var_long)) %>%
  group_by(group, var_long, axis) %>%
  summarize(axis_loading = mean(axis_loading)) %>%
  ungroup() %>%
  complete(group, var_long, axis, fill = list(axis_loading = NA)) %>%
  mutate(inout = ifelse(endsWith(var_long, "DHI"), "Response", "Predictor"), 
         group = fct_relevel(group, "British Columbia", 
                             "Broadleaf", 
                             "Coniferous", 
                             "Mixed Wood", 
                             "Wetland-Treed", 
                             after = 0),
         var_long = fct_relevel(var_long, 
                                c("Canopy Height", "Canopy Cover", "Structural Complexity",
                                "Basal Area", "Aboveground Biomass", "Gross Stem Volume",
                                "Cumulative DHI", "Variation DHI", "Minimum DHI")))

# nojitter/lines
parcoord_noflip <- data %>%
  ggplot(aes(x = var_long, y = axis_loading)) +
  #geom_boxplot(col = "#00000025", outlier.alpha = 0) +

  geom_line(aes(col = group, group = group), lty = "dashed") +
  geom_point(aes(col = group, group = group, shape = group, fill = group)) +
  # geom_hline(yintercept = 0, lty = "dashed", col = "red") +
  facet_wrap(~paste(axis, inout),
             scale = "free") +
  #theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(x = NULL,
       y = "Axis Loading (absolute)",
       col = NULL,
       shape = NULL,
       fill = NULL) +
  # coord_flip() +
  scale_colour_discrete_c4a_cat(palette = "bold") +
  scale_fill_discrete_c4a_cat(palette = "bold") +
  scale_shape_manual(values = c(21, 22, 22, 22, 22, 25, 25, 25, 25, 25))

ggsave(here::here("outputs", "parcoord_noflip.svg"), plot = parcoord_noflip,
       height = 6, width = 8)
```

# non normalized
loadings_rgb <- clean_loads_abs %>% 
  group_by(axis) %>%
  pivot_wider(names_from = var,
              values_from = axis_loading) %>%
  ungroup() %>%
  replace(is.na(.), 0) %>%
  mutate(across(elev_cv:elev_p95, \(x) ifelse(x > 0.5, 0.5, x))) %>%
  mutate(dhi_rgb = rgb(CumDHI, VarDHI, MinDHI),
         pri_rgb = rgb(elev_p95, percentage_first_returns_above_2m, elev_cv, maxColorValue = 1),
         mod_rgb = rgb(basal_area, total_biomass, gross_stem_volume, maxColorValue = 1)) %>%
  select(strata, axis, percentage_first_returns_above_2m, elev_p95, elev_cv, ends_with("rgb")) %>%
  pivot_longer(ends_with("rgb"))

# normalized to max
loadings_rgb <- clean_loads %>% 
  group_by(var) %>%
  mutate(axis_loading = axis_loading / max(axis_loading)) %>%
  group_by(axis) %>%
  pivot_wider(names_from = var,
              values_from = axis_loading) %>%
  ungroup() %>%
  replace(is.na(.), 0) %>%
  #mutate(across(elev_cv:elev_p95, \(x) ifelse(x > 0.5, 0.5, x))) %>%
  mutate(dhi_rgb = rgb(CumDHI, VarDHI, MinDHI),
         pri_rgb = rgb(elev_p95, percentage_first_returns_above_2m, elev_cv, maxColorValue = 1),
         mod_rgb = rgb(basal_area, total_biomass, gross_stem_volume, maxColorValue = 1)) %>%
  #select(strata, axis, percentage_first_returns_above_2m, elev_p95, elev_cv, ends_with("rgb")) %>%
  pivot_longer(ends_with("rgb"))

becs <- bec_normal %>%
  pull(bec) %>%
  unique()

rgb_loads_split <- loadings_rgb %>%
  filter(strata %in% becs) %>%
  group_by(axis, name) %>%
  group_split()
```

labels = c("DHI", "B - Modelled RDA1", "A - Primary RDA1", "DHI", "D - Modelled RDA2", "C - Primary RDA2")

# for (i in 1:length(rgb_maps)) {
#   print(i)
#   gg <- rgb_maps[[i]] + labs(title = labels[[i]])
#   
#   ggsave(here::here("outputs", "black_plots", paste(i, ".png")),
#          plot = gg,
#          height = 3,
#          width = 3)
# }

ggsave(here::here("outputs", "test_black.png"), rgb_maps[[6]])

patchwork_rgbplots <- (rgb_maps[[3]] + rgb_maps[[2]]) / 
                      (rgb_maps[[6]] + rgb_maps[[5]]) &
  plot_annotation(tag_levels = "A")

# ggsave(here::here("outputs", "rgbplots_norm.png"), patchwork_rgbplots,
#        height = 6,
#        width = 8)
# 
# print("png saved")

# ggsave(here::here("outputs", "rgbplots_norm.pdf"), patchwork_rgbplots,
#        height = 6,
#        width = 8)
# 
# print("pdf saved")
```

# parallel coordinate pos neg

```{r}
neg_pos <- clean_loads_raw %>%
  left_join(keys$bec %>%
              select(strata = zone, group)) %>%
  left_join(keys$continuous %>%
              select(var = variable, var_long)) %>%
  mutate(group = ifelse(is.na(group), strata, group),
         var_long = ifelse(is.na(var_long), var, var_long)) %>%
  group_by(group, var_long, axis) %>%
  summarize(n_pos = sum(ifelse(axis_loading > 0, 1, 0)),
            n_neg = sum(ifelse(axis_loading < 0, 1, 0)),
            axis_loading = mean(axis_loading)
            ) %>%
  ungroup() %>%
  complete(group, var_long, axis, fill = list(axis_loading = NA,
                                              n_pos = NA,
                                              n_neg = NA)) %>%
  mutate(pos_neg = case_when(n_pos > 0 & n_neg > 0 ~ "Mixed",
                             n_pos > 0 ~ "Positive",
                             n_neg > 0 ~ "Negative")) %>%
  select(-n_pos, -n_neg, -axis_loading)

>>>>

```{r packages}
library(vegan)
library(tidyverse)
library(cols4all)
library(terra)
library(sf)
library(patchwork)

source(here::here("scripts", "get_keys.R"))
```

@@

```{r data-prep}
zones <- keys$bec %>% pull(zone)
groups <- keys$bec %>% pull(group) %>% unique()

bec_sample_loc <- here::here("data", "samples", "vlce_extracted.shp")

@@

bec_normal <- bec_df %>%
  mutate(across(basal_area:VarDHI,  .fns = ~ normal(.x))) 
```

set.seed(69420)

strata_nested <- bind_rows(bec_nested, class_nested) %>%
  mutate(data = map(data, slice_sample, n = 500))
# this slice sample is the ANALYTICAL difference between 4_vegan and 
# 4_vegan_clean, other than the actual cleanup of the .rmd file

global <- strata_nested %>%
  filter(strata %in% zones) %>%
  unnest(data)
```

```{r global-model}
forest_types <- global %>%
  pull(category) %>%
  unique()

dhi_tab <- global %>%
  select(ends_with("DHI"))

primary <- global %>%
  select(elev_p95,
         elev_cv,
         percentage_first_returns_above_2m)

modelled <- global %>%
  select(basal_area,
         total_biomass,
         gross_stem_volume)

rda_out <-
  rda(
    dhi_tab ~ elev_p95 + percentage_first_returns_above_2m + elev_cv + # simple/raw/primary
      basal_area + gross_stem_volume + total_biomass,
    # modelled
    data = global
  )

@@

rcorr <- summary(rda_out)$sites %>%
  as.data.frame() %>%
  select(starts_with("RDA")) %>%
  bind_cols(global) %>%
  select(-strata,-category) %>%
  as.matrix() %>%
  Hmisc::rcorr()

@@

all_data_loadings <- left_join(rvals, pvals) %>%
  filter(axis != "RDA3") %>%
  mutate(sig = cut(
    axis_p,
    breaks = c(0, 0.001, 0.01, 0.05, .10, 1),
    labels = c("***", "**", "*", ".", "NS"),
    right = F
  ))

@@

@@

@@

@@

bec_normal %>%
  pivot_longer(cols = ends_with("DHI"),
               names_to = "dhi",
               values_to = "dhi_val") %>%
  pivot_longer(
    cols = basal_area:total_biomass,
    names_to = "struct",
    values_to = "struct_val"
  ) %>%
  ggplot(aes(x = struct_val, y = dhi_val)) +
  geom_hex() +
  facet_grid(rows = vars(dhi),
             cols = vars(struct),
             scales = "free") +
  labs(x = "Structure Value (Normalized)",
       y = "DHI Value (Normalized)")
```

@@

```{r fracts-out}
fracts_out <- outs %>%
  mutate(indfracts = map(var, .f = function(x) {
    x$part$indfract %>%
      mutate(type = c("primary", "modelled", "overlap", "residual")) %>%
      select(type, Adj.R.squared) %>%
      pivot_wider(names_from = type, values_from = Adj.R.squared)
    }
    )) %>%
  unnest(indfracts) %>%
  arrange(desc(residual)) %>%
  select(strata, primary, modelled, overlap) %>%
  pivot_longer(c(primary, modelled, overlap)) %>%
  mutate(value = ifelse(value > 0, value, 0)) %>%
  pivot_wider()

@@

Fracts Bar Plot (Figure)

@@

@@

@@

@@

```{r varexp}
rda_varexp <- outs %>%
  mutate(varexp = map(rda, .f = \(rda) {
    rda_sum <- summary(rda)
    var_exp <- rda_sum$constr.chi / rda_sum$tot.chi
  })) %>%
  unnest(varexp) %>%
  arrange(desc(varexp)) %>%
  select(strata, varexp)

rda_varexp %>%
  arrange(strata %in% forest_types,
          desc(varexp))
```

@@

sig_loads %>%
  filter(anova_sig != "NS") %>%
  filter(endsWith(var, "DHI"),
         axis == "RDA2") %>%
  arrange(strata) %>%
  select(strata, var, axis_loading) %>%
  pivot_wider(names_from = var,
              values_from = axis_loading)
# when a second axis is present, it has a weak loading on the cumulative DHI, as compared to the other DHIs
# while the majority of the variability is contained within all three DHIs together, the second axis, when present, is more linked to the minmimum and variation DHI.

@@

@@

  write_lines(strata_drawio, here::here(drawio_dir, paste0(s, ".drawio")))
}

file.copy(here::here("outputs", "drawio", "British Columbia.drawio"), here::here("outputs", "British Columbia.drawio"), overwrite = T)
```

Climate Table (Table 2)

bec_tab <- keys$bec %>%
  left_join(bec_clim, by = c("zone" = "bec")) %>%
  select(-order) %>%
  arrange(group, zone) %>%
  relocate(group) %>%
  mutate(across(ppt:tmin, \(x) round(x, digits = 1))) %>%
  left_join(strata_nested %>% unnest(data) %>% count(strata),
            by = c("zone" = "strata"))

# normalized to max
loadings_rgb <- clean_loads_abs %>% 
  group_by(var) %>%
  mutate(axis_loading = axis_loading / max(axis_loading)) %>%
  group_by(axis) %>%
  pivot_wider(names_from = var,
              values_from = axis_loading) %>%
  ungroup() %>%
  replace(is.na(.), 0) %>%
  #mutate(across(elev_cv:elev_p95, \(x) ifelse(x > 0.5, 0.5, x))) %>%
  mutate(dhi_rgb = rgb(CumDHI, VarDHI, MinDHI),
         pri_rgb = rgb(elev_p95, percentage_first_returns_above_2m, elev_cv, maxColorValue = 1),
         mod_rgb = rgb(basal_area, total_biomass, gross_stem_volume, maxColorValue = 1)) %>%
  #select(strata, axis, percentage_first_returns_above_2m, elev_p95, elev_cv, ends_with("rgb")) %>%
  pivot_longer(ends_with("rgb"))

rgb_loads_split <- loadings_rgb %>%
  filter(strata %in% zones) %>%
  group_by(axis, name) %>%
  group_split()
```

```{r}
flip_cum <- function(data) {
  if(data %>% filter(var == "CumDHI") %>% nrow() == 0) {
    return(data)
  }
  flip <- data %>%
    filter(var == "CumDHI") %>%
    pull(axis_loading) %>%
    sign()

Parcoord Noflip (Figure)

```{r}
data_parcoord <- clean_loads_joined %>%
  # mutate(axis_loading = abs(axis_loading)) %>%
  group_by(group, var_long, axis) %>%
  summarize(axis_loading = mean(axis_loading)) %>%
  ungroup() %>%
  complete(group, var_long, axis, fill = list(axis_loading = NA)) %>%
  mutate(inout = ifelse(endsWith(var_long, "DHI"), "Response", "Predictor"), 
         group = fct_relevel(group, "British Columbia", 
                             "Broadleaf", 
                             "Coniferous", 
                             "Mixed Wood", 
                             "Wetland-Treed", 
                             after = 0),
         var_long = ifelse(endsWith(var_long, "DHI"), var_long, str_replace_all(var_long, " ", "\n")),
         var_long = fct_relevel(var_long, 
                                c("Structural\nComplexity", "Canopy\nHeight", "Canopy\nCover",
                                "Basal\nArea", "Aboveground\nBiomass", "Gross\nStem\nVolume",
                                "Cumulative DHI", "Variation DHI", "Minimum DHI")))

# nojitter/lines
parcoord_noflip <- data_parcoord %>%
  ggplot(aes(x = var_long, y = axis_loading)) +
  #geom_boxplot(col = "#00000025", outlier.alpha = 0) +

# meg scatterplots

```{r}
scatter <- bec_normal %>%
  mutate(label = case_when(bec %in% c("CWH", "MS") ~ bec,
                           T ~ "Other"))

Supplemental information Hexplot template

IQR sorted plot of canopy cover

```{r iqr-plot}
strata_nested %>% 
  unnest(data) %>% 
  ggplot(aes(x = fct_reorder(strata, percentage_first_returns_above_2m, IQR), y = percentage_first_returns_above_2m)) + 
  geom_boxplot() + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Study Area Map (Figure 1)

@@
